\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-11 pldi/draft.tex
Comments on dangling fifos and zombie producers

To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts. % to their parallel equivalents.
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.

@@

@@

\begin{lstlisting}[float=h, numbers=none]
  mkfifo t1 t2
  grep "foo" f1 > t1 &
  grep "foo" f2 > t2 &
  cat t1 t2
\end{lstlisting}

@@

\kk{Maybe say something about the rest of the transformations: (i)
  changing a command with multiple inputs to a cat followed by the
  command, (ii) Adding an identity node in the place of two edges, and
  (iii) adding a split after a command.}

@@

This section describes challenges related to the implementation of
\sys. It starts by describing how the \sys front-end translates a
shell script to the dataflow graph model defined in \S\ref{ir}, by
applying transformation on parallelizable regions~\sx{front-end}.  It
closes with the \sys optimizer, which optimizes the dataflow graph
exposing data-parallelism, and outputs an optimized
parallel script~\sx{optimizer}.

@@

\subsection{Optimizer}
\label{optimizer}

\kk{This subsection has to only talk about the cool shell constructs
  that are added and about optimizations that are not in the dataflow
  graph (like file splitting), and the implementation of split, the
  pure merges , etc...}

@@

Given a dataflow graph representing a shell program, the \sys optimizer outputs an optimized dataflow graph in which data parallelism opportunities have been exploited.
More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges.

The optimizer also applies custom transformations on several commands in \pur
to further improve their performance. Two notable examples are the
transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
that have more than one input are tranformed to \ttt{sort} commands on
each input, followed by a \ttt{sort -m}, to produce the final sorted
result. The transformation on occurences of \ttt{wc} that have more
than input \ttt{i1, t2} is illustrated below as a shell script:
% \nv{do not understand i1, t2}

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
  paste -d '+' \
   <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
   <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
   bc | tr -s '\n' ' ' |
   sed 's/^/   /' | sed 's/$/\ /'
\end{lstlisting}

The above script takes the outputs produced by two \ttt{wc} commands
and aggregates them accordingly.

@@

\subsection{Implementation Nuggets}

@@

% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
% \ttt{SIGPIPE} 
However, if its consumer exits early, the producer needs to be notified so that it stops appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.

In \unix, this is achieved by an out-of-band error mechanism:
  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
}
primarily because pipes and pipelines are at the heart of its philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:
\begin{lstlisting}[language=sh, numbers=none]
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

@@

>>>>

To address this challenge, we develop \sys, a shell parallelizing\todo{/that parallelizes} POSIX shell scripts.
\kk{Mention that it outputs shell scripts.} % to their parallel equivalents.
% \sys offers both classes of users in the second group such much-needed automation.
\sys benefits both programmer groups mentioned earlier, with a particular emphasis on users.
Command developers are provided a set of abstractions, akin to lightweight type annotations, for expressing the parallelizability properties of their commands:
  rather than providing a command's full observable behavior, these annotations focus primarily on their interaction with state.
Shell users, on the other hand, are provided full automation:
  \sys extracts parallelism latent in their scripts by analyzing their parallelizability properties. % of their fragments.
% benefiting scripting experts and non-experts alike by offering automated POSIX-aware parallelization of shell-scripts as long as command developers
\sys's transformations are conservative, in that they do not attempt to parallelize fragments that lack sufficient information---\ie at worst, \sys will choose to not improve performance rather than risking breakage.

@@

\kk{Here we should reiterate on the fact that \sys takes a script and outputs a script.}

@@

\begin{lstlisting}[language=sh, numbers=none]
  mkfifo t1 t2
  grep "foo" f1 > t1 &
  grep "foo" f2 > t2 &
  cat t1 t2
\end{lstlisting}

@@

\kk{I think we should say something about the rest of the
  transformations: (i) changing a command with multiple inputs to a
  cat followed by the command to enable the previous optimization,
  (ii) adding a split node after a command (without saying anything
  about its implementation for now, and (iii) Adding an identity node
  in the place of an edge.}

@@

This section describes challenges related to the implementation of
\sys. It starts by describing how the \sys front-end translates a
shell script to the dataflow graph model defined in \S\ref{ir}, by
applying transformation on parallelizable regions~\sx{front-end}.  It
closes with the \sys runtime\todo{/backend} and the technical
challenges that were addressed to achieve performance
benefits~\sx{optimizer}.

@@

\subsection{Runtime/Back End}
\label{optimizer}

@@

Given a dataflow graph representing a shell program, the \sys
optimizer outputs an optimized dataflow graph in which data
parallelism opportunities have been exploited.  More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}.
%% If the incoming edges of the node are less than the desirable graph
%% width, it performs edge splits, and then it performs maximal node
%% splits.
The final graph can then be implemented by spawning a process for each
node, and redirecting the inputs and outputs according to the graph
edges.

@@

%% \subsection{Implementation Nuggets}

@@

% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a producer exits after it has produced and
sent all of its output to a pipe\todo{/its output channel}.
% \ttt{SIGPIPE} 
However, if its consumer exits early, the producer needs to be
notified so that it stops \todo{appending to the stream/trying to push
  data to its output channel}.
% ---so that it avoids unnecessary work and buffering challenges in the kernel.
In \unix, this is achieved by an out-of-band error mechanism:
  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
}
primarily because pipes and pipelines are at the heart of its philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes\kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:\kk{Proposed alternative: ``Consider the following shell script:''}
\begin{lstlisting}[language=sh, numbers=none]
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

@@

\paragraph{Pure Fold Function Implementations}

\kk{Maybe say something about some of the pure fold functions that we
  implemented. Interesting examples include sort and wc.}

%% The optimizer also applies custom transformations on several commands in \pur
%% to further improve their performance. Two notable examples are the
%% transformations on \ttt{sort} and \ttt{wc}. Occurences of \ttt{sort}
%% that have more than one input are tranformed to \ttt{sort} commands on
%% each input, followed by a \ttt{sort -m}, to produce the final sorted
%% result. The transformation on occurences of \ttt{wc} that have more
%% than input \ttt{i1, t2} is illustrated below as a shell script:
%% % \nv{do not understand i1, t2}

%% \begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
%%   paste -d '+' \
%%    <(wc i1 | tr -s ' ' '\n' | tail -n +2) \
%%    <(wc i2 | tr -s ' ' '\n' | tail -n +2) |
%%    bc | tr -s '\n' ' ' |
%%    sed 's/^/   /' | sed 's/$/\ /'
%% \end{lstlisting}

%% The above script takes the outputs produced by two \ttt{wc} commands
%% and aggregates them accordingly.
